Mahotas: Open source software for scriptable computer vision

### 2013-01-05 paper.tex
MIN Update date
TXT Corrections from a read through

\title{Mahotas: Open source software for scriptable computer vision}
\author{Luis Pedro Coelho\\
Lane Center for Computational Biology, Carnegie Mellon University\\
Instituto de Medicina Molecular}
\date{September 2012}

@@

The interface is in Python, a dynamic programming language, which is very
appropriate for fast development, but the algorithms are implemented in \cpp{}
and are tuned for speed.

@@

Additionally, by integrating into the Python numeric ecosystem, users can use
other packages in a seamless way. In particular, mahotas does not implement any
machine learning functionality, but rather advises the user to use another,
specialised package, such as scikits-learn or milk.

@@

\begin{description}
\item[\textsc{surf}] Speeded-up Robust Features~\citep{eth_biwi_00517}. This
includes both keypoint detection and descriptor computation.
\item[features] Global feature descriptors. In particular, Haralick texture
features, Zernike moments, local binary patterns, and threshold adjacency
statistics (both the original~\citep{Hamilton2007} and the parameter-free
versions~\citep{Coelho2010}).
\item[wavelet] Haar and Daubechies wavelets. Forward and inverse transforms are
supported.
\item[morphological functions] Erosion and dilation, as well as some more
complex operations built on these. There are both binary and grayscale
implementations of these operators.
\item[watershed] seeded watershed and distance map
transforms~\citep{felzenszwalb}.
\item[filtering] Gaussian filtering, edge finding, and general convolutions.
\item[polygon operations] convex hull, polygon drawing.
\end{description}

In general, functions operate on all data types. This is performed without any
extra memory copies. Mahotas is heavily optimised for both speed and memory
usage (it can be used with very large arrays).

@@

Code for this and other examples is present in the mahotas source distribution
under the \texttt{demos/} directory. In this example, we load an image and find
SURF interest point and descriptors on it.

@@

Using numpy operations and milk, we can select only the descriptors and cluster
them into a group of five:

\begin{python}
descrs = spoints[:,6:]
values, _ = milk.kmeans(descrs, 5)
\end{python}

@@

Mahotas is mostly written in \cpp, but this is completely hidden from the user
as there are hand-written Python wrappers for all functions (automatically
generated wrappers inevitably lead to worse error messages and are less
flexible).

@@

So, for example, this is how \emph{erode} is implemented.\footnote{This is the
generic version of the code. Erode, like a few other functions, has two
versions, a fast version, limited to two-dimensional, contiguous, images and a
generic one presented here.} \code{py\_erode} consists mostly of boiler-plate
code:

@@

This functions retrieves the arguments, performs some sanity checks, performs a
bit of initialisation, and finally, switches in the input type with the help of
the \code{SAFE\_SWITCH\_ON\_INTEGER\_TYPES} macro, which call the right
specialisation of the template that does the actual work. In this example
\code{erode} implements erosion:

@@

The template machinery makes the functions that use it very simple and easy to
read. The only downside is that there is some expansion of code size when the
compiler instantiates the function for the several integer and floating point
types. Given the small size of these functions, the total size of the compiled
library is reasonable (circa 6MiB on an Intel-based 64~bit system).

@@

\begin{table}
\centering
\begin{tabular}{lrrrrr}
\toprule
Operation & mahotas & pymorph & scikits-image & OpenCV\\
\midrule
erode                &      1.6 &     15.1 &      7.4 &      0.4 & \\
dilate               &      1.5 &      9.1 &      7.3 &      0.4 & \\
open                 &      3.2 &     24.3 &     14.8 &       NA & \\
median filter (2)    &    226.9 &       NA &   2034.0 &       NA & \\
median filter (10)   &   2810.9 &       NA &   1877.1 &       NA & \\
center mass          &      5.0 &       NA &   3611.2 &       NA & \\
sobel                &     34.1 &       NA &     62.5 &      6.2 & \\
cwatershed           &    174.8 &  58440.3 &    287.3 &     44.9 & \\
daubechies           &     18.8 &       NA &       NA &       NA & \\
haralick             &    233.1 &       NA &   7760.7 &       NA & \\
\bottomrule
\end{tabular}
\caption{Efficiency Results for mahotas, pymorph, scikits-image, and openCV
(through Python wrappers). Shown are values as multiples of the time that
\code{numpy.max(image)} takes to compute the maximum pixel value in the image
(all operations are over the same image). For scikits-image, features on the
grey-scale cooccurrence matrix were used instead of Haralick features, which it
does not support. In the case of \emph{median filter}, the radius of the
structuring element is shown in parentheses.}
\label{tab:efficiency}
\end{table}

Table~\ref{tab:efficiency} shows timings for different operations. These were
normalized to multiples of the time it takes to go over the image and find its
maximum pixel value (this was done using \code{numpy.max(image)}). The
measurements shown were obtained on an Intel 64~bit system, running Ubuntu
Linux. However, due to the normalization, measurements obtained on another
system (Intel 32~bits running Mac OS) were qualitatively similar (data not
shown).

@@

OpenCV is the fastest library, but this comes at the cost of some flexibility.
Arguments to its functions must be of the exact expected type and it is
possible to crash the interpreter if types do match the expected type (in the
other libraries, including mahotas, all types are checked and an exception is
generated which can be caught by user code). This is particularly relevant for
interactive use as the user is often exploring and is willing to pay the speed
cost of a few extra type conversions to avoid a hard-crash.

Among the other libraries, mahotas is the fastest. Pymorph, even though it is
implemented in Python only, intelligently uses arithmetic operations for this
operation and can be very fast. However, for more complex methods, such as
watershed; its pure Python approach is very inefficient. The one exception is
that median filtering with a large structuring element is faster in
scikit-image. In fact, that library uses an algorithm with better asymptotic
behavior for this operation.

\bigskip

@@

Mahotas does not include machine learning related functionality, such as
$k$-means clustering or classification methods. This is the result of an
explicit design decision. Specialised machine learning packages for Python
already
exist~\citep{Pedregosa:2011:SML:2078183.2078195,springerlink:10.1007/978-3-540-30116-5_58,Schaul:2010:PYB:1756006.1756030,Sonnenburg:2010:SML:1756006.1859911}.
A good classification (or other) system can benefit both computer vision users
and others. As these projects all use Numpy arrays as their data types, it is
easy to use functionality from the different project seamlessly (no copying of
data is necessary).

Python is an ideal language for fast development of both applications and
scientific software. For this platform, Mahotas is a fast library of computer
vision and image processing functions. It is implemented in \cpp{}, as the
standard Python interpreter is too slow for a direct Python implementation.
However, all of the Python interface code is hand-written, as opposed to using
automatic interface generators like Swig~\citep{Beazley2003599}. This is more
work intially, but the end result is of much higher quality, especially when it
comes to giving useful error messages (e.g., when a type mismatch occurs, an
automatic system will often be forced to resort to a generic message as it does
not have any knowledge of what the arguments mean besides their automatically
inferred types).

@@

Mahotas includes code ported and incorporated from other projects. Initially,
it was used in reproducing the functionality in the Subcellular Location Image
Classifier (SLIC) tool from Robert F.\ Murphy's
Lab~\citep{springerlink:10.1007/978-0-387-45524-2_47} and the initial versions
of mahotas were designed explicitly to support that functionality. The
\textsc{surf} implementation is a port from the code from
\textit{dlib},\footnote{Dlib's webpage is at \url{http://dlib.net}.} a very
good \cpp{} library by Davis King. I also gleaned some insight into the
implementation of these features from Christopher Evan's OpenSURF library and
its documentation \citep{evans2009}.\footnote{OpenSURF is available at
\url{http://www.chrisevansdev.com/computer-vision-opensurf.html}, where several
documents describe details of the implementation.} The code which interfaces
with the FreeImage library, was written by Zachary Pincus and some of the
support code was written by Peter J. Verveer for the \code{scipy.ndimage}
project. All of these contributions were integrated while respecting the
software licenses under which the original code had been released. Robert Webb,
a summer student at Carnegie Mellon University, worked with me on the initial
local binary patterns implementation. Finally, I thank the several users who
have reported bugs, submitted small fixes, and participated on the project
mailing list.

Stéfan van der Walt and Andreas Müller offered helpful comments on a draft
version of this manuscript.

>>>>

\title{Mahotas: Open source software for scriptable computer vision}
\author{Luis Pedro Coelho\\
Lane Center for Computational Biology, Carnegie Mellon University\\
Instituto de Medicina Molecular}
\date{January 2013}

@@

The interface is in Python, a dynamic programming language, which is very
appropriate for fast development, but the algorithms are implemented in \cpp{}
and are tuned for speed. The library is designed to fit in with the scientific
software ecosystem in this language and can leverage the existing
infrastructure developed in that language.

@@

It contains over 100~functions with functionality ranging from traditional
image filtering and morphological operations to more modern wavelet
decompositions and local feature computations. Additionally, by integrating
into the Python numeric ecosystem, users can use other packages in a seamless
way. In particular, mahotas does not implement any machine learning
functionality, but rather advises the user to use another, specialised package,
such as scikits-learn or milk.

@@

\begin{description}
\item[\textsc{Surf}] Speeded-up Robust Features~\citep{eth_biwi_00517}. This
includes both keypoint detection and descriptor computation.
\item[Features] Global feature descriptors. In particular, Haralick texture
features, Zernike moments, local binary patterns, and threshold adjacency
statistics (both the original~\citep{Hamilton2007} and the parameter-free
versions~\citep{Coelho2010}).
\item[Wavelet] Haar and Daubechies wavelets. Forward and inverse transforms are
supported.
\item[Morphological functions] Erosion and dilation, as well as some more
complex operations built on these. There are both binary and grayscale
implementations of these operators.
\item[Watershed] seeded watershed and distance map
transforms~\citep{felzenszwalb}.
\item[Filtering] Gaussian filtering, edge finding, and general convolutions.
\item[Polygon operations] convex hull, polygon drawing.
\end{description}

Numpy arrays contain data of a specific type, such \code{unsigned 8 bit
integer} or floating point numbers. While natural colour images are typically
8~bits, scientific data is often larger and processing can result in floating
point images. Mahotas works on all datatypes. This is performed without any
extra memory copies. Mahotas is heavily optimised for both speed and memory
usage (it can be used with very large arrays).

@@

Code for this and other examples is present in the mahotas source distribution
under the \texttt{demos/} directory. In this example, we load an image, find
SURF interest points, and compute descriptors.

@@

The \code{surf.surf} function returns both the descriptors and their meta data.
We use numpy operations to retain only the descriptors (the meta data is in the
first five positions):

\begin{python}
descrs = spoints[:,6:]
\end{python}

Using milk, we cluster the descriptors into five groups:

\begin{python}
values, _ = milk.kmeans(descrs, 5)
\end{python}

@@

Mahotas is mostly written in \cpp, but this is completely hidden from the user
as there are hand-written Python wrappers for all functions. Automatically
generated wrappers inevitably lead to worse error messages and are less
flexible.

@@

So, for example, this is how \emph{erode} is implemented.\footnote{This is the
generic version of the code. Erode, like a few other functions, has two
versions, a fast version, limited to two-dimensional, contiguous, images and a
generic one presented here. The selection between the two implementations is
done automatically for the user.} \code{py\_erode} consists mostly of
boiler-plate code:

@@

This functions retrieves the arguments, performs some sanity checks, performs a
bit of initialisation, and finally, switches in the input type with the help of
the \code{SAFE\_\-SWITCH\_\-ON\_\-INTEGER\_\-TYPES} macro, which call the right
specialisation of the template that does the actual work. In this example
\code{erode} implements erosion:

@@

The template machinery makes the functions that use it very simple and easy to
read. The only downside is that there is some expansion of code size when the
compiler instantiates the function for the several integer and floating point
types. Given the small size of these functions, the total size of the compiled
library is reasonable (circa 6MiB on an Intel-based 64~bit system for the whole
library).

@@

\begin{table}
\centering
\begin{tabular}{lrrrrrr}
\toprule
Operation            &  mahotas &  pymorph & scikits-image & OpenCV \\
\midrule
erode                &      1.6 &     15.1 &      7.4 &      0.4  \\
dilate               &      1.5 &      9.1 &      7.3 &      0.4  \\
open                 &      3.2 &     24.3 &     14.8 &       NA  \\
median filter (2)    &    226.9 &       NA &   2034.0 &       NA  \\
median filter (10)   &   2810.9 &       NA &   1877.1 &       NA  \\
center mass          &      5.0 &       NA &   3611.2 &       NA  \\
sobel                &     34.1 &       NA &     62.5 &      6.2  \\
cwatershed           &    174.8 &  58440.3 &    287.3 &     44.9  \\
daubechies           &     18.8 &       NA &       NA &       NA  \\
haralick             &    233.1 &       NA &   7760.7 &       NA  \\
\bottomrule
\end{tabular}
\caption{Efficiency Results for mahotas, pymorph, scikits-image, and openCV
(through Python wrappers). Shown are values as multiples of the time that
\code{numpy.max(image)} takes to compute the maximum pixel value in the image
(all operations are over the same image). For scikits-image, features on the
grey-scale cooccurrence matrix were used instead of Haralick features, which it
does not support. In the case of \emph{median filter}, the radius of the
structuring element is shown in parentheses. NA stands for ``Not Available.''}
\label{tab:efficiency} \end{table}

Table~\ref{tab:efficiency} shows timings for different operations. These were
normalized to multiples of the time it takes to go over the image and find its
maximum pixel value (this was done using \code{numpy.max(image)}). The
measurements shown were obtained on an Intel 64~bit system, running Ubuntu
Linux. However, due to the normalization, measurements obtained on another
system (Intel 32~bits running Mac OS) were qualitatively similar.

@@

OpenCV is the fastest library, but this comes at the cost of some flexibility.
Arguments to its functions must be of the exact expected type and it is
possible to crash the interpreter if types do match the expected type (in the
other libraries, including mahotas, all types are checked and an exception is
generated which can be caught by user code). This is particularly relevant for
interactive use as the user is often exploring and is willing to pay the speed
cost of a few extra type checks and conversions to avoid a hard-crash.

Among the other libraries, mahotas is the fastest. Pymorph, even though it is
implemented in Python only, intelligently uses arithmetic operations for
morphological operation and can be very fast. However, for more complex
methods, such as watershed; its pure Python approach is very inefficient. The
one exception is that median filtering with a large structuring element is
faster in scikit-image. In fact, that library uses an algorithm with better
asymptotic behavior for this operation.

\subsection{Distribution and Installation}

@@

Mahotas does not include machine learning related functionality, such as
$k$-means clustering or classification methods. This is the result of an
explicit design decision. Specialised machine learning packages for Python
already
exist~\citep{Pedregosa:2011:SML:2078183.2078195,springerlink:10.1007/978-3-540-30116-5_58,Schaul:2010:PYB:1756006.1756030,Sonnenburg:2010:SML:1756006.1859911}.
A good classification system can benefit both computer vision users
and others. As these projects all use Numpy arrays as their data types, it is
easy to use functionality from the different project seamlessly (no copying of
data is necessary).

Mahotas is implemented in \cpp{}, as the standard Python interpreter is too
slow for a direct Python implementation. However, all of the Python interface
code is hand-written, as opposed to using automatic interface generators like
Swig~\citep{Beazley2003599}. This is more work initially, but the end result is
of much higher quality, especially when it comes to giving useful error
messages (e.g., when a type mismatch occurs, an automatic system will often be
forced to resort to a generic message as it does not have any knowledge of what
the arguments mean besides their automatically inferred types).

@@

Mahotas includes code ported and incorporated from other projects. Initially,
it was used in reproducing the functionality in the Subcellular Location Image
Classifier (SLIC) tool from Robert F.\ Murphy's
Lab~\citep{springerlink:10.1007/978-0-387-45524-2_47} and the initial versions
of mahotas were designed explicitly to support that functionality. The
\textsc{surf} implementation is a port from the code from
\textit{dlib},\footnote{Dlib's webpage is at \url{http://dlib.net}.} a very
good \cpp{} library by Davis King. I also gleaned some insight into the
implementation of these features from Christopher Evan's OpenSURF library and
its documentation \citep{evans2009}.\footnote{OpenSURF is available at
\url{http://www.chrisevansdev.com/computer-vision-opensurf.html}, where several
documents describe details of the implementation.} The code which interfaces
with the FreeImage library, was written by Zachary Pincus and some of the
support code was written by Peter J. Verveer for the \code{scipy.ndimage}
project. All of these contributions were integrated while respecting the
software licenses under which the original code had been released. Robert Webb,
a summer student at Carnegie Mellon University, worked with me on the initial
local binary patterns implementation. Finally, I thank the several users who
have reported bugs, submitted fixes, and participated on the project
mailing list.

St\'efan van der Walt and Andreas M\"uller offered helpful comments on a draft
version of this manuscript.
